D3064R0: Bringing the Laws of Physics, Computer-System Constraints, and Compilers to Bear on the OOTA Problem

### 2024-02-04 ootavacuous.tex
FYI, possible progress on language-level OOTA problem
More additions: the section on executions.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

\section{OOTA, Semantic Dependencies, and Executions}
\label{sec:OOTA; Semantic Dependencies; and Executions}

@@

\subsection{What Exactly is an Execution?}
\label{sec:What Exactly is an Execution?}

@@

>>>>

\section{OOTA and Semantic Dependencies}
\label{sec:OOTA and Semantic Dependencies}

@@

\section{What is an Execution?}
\label{sec:What is an Execution?}

@@

\subsection{Abstract Executions}
\label{sec:Abstract Executions}

The C++ standard describes the execution of a program in terms of
``a parameterized nondeterministic abstract machine'' in 4.1.2p1
(\co{intro.abstract}).
This description specifies how the abstract machine carries out the
operations of a source program in great, but not complete, detail:
\begin{itemize}
\item	Some of the abstract machine's characteristics are
	implementation defined, including things like the number of
	bits in the various integer types
	or whether the \co{char} type is signed.
\item	Some aspects of an execution are unspecified or nondeterministic,
	including things like the order of evaluation of the operands
	of most binary operators or of the arguments in a function call.
	Implementations may choose from a set of allowed behaviors.
\item	Some actions are deemed to have undefined results; the standard
	says essentially nothing about programs that can give rise to
	undefined behavior.
\item	Asynchronous actions (i.e., signal handlers) are largely ignored.
\item	Input and output are not described in any detail.
\end{itemize}
In addition to these points, the standard does not specify which store
an atomic load must read from, beyond requiring that the overall
pattern of loads and stores be consistent with the C++ memory model.
We assume that programs will not indulge in any other computations that
could be expected to vary spontaneously from one execution to another,
such as basing a dependency on the time of day or a process ID.

The implementation-defined aspects can affect whether or not an
abstract execution contains a semantic dependency.
As an example consider the following, where the type of \co{c} is \co{char}:
\begin{quote}
\begin{verbatim}
y = x * (c >= 0);
\end{verbatim}
\end{quote}
Here \co{y} is semantically dependent on \co{c} in executions
for which \co{char} is a signed type,
but not those for which it is unsigned.

The same is theoretically true for the nondeterministic aspects of an
execution.
Consider this example, with \co{i} initially zero:
\begin{quote}
\begin{verbatim}
r1 = ++i / ++i;
x = r1 * z;
\end{verbatim}
\end{quote}
Because the order of evaluation of the division's arguments is undetermined,
the value calculated for \co{r1} might be zero ($1/2$ truncated) or two
($2/1$).
In the former case there is no semantic dependency from \co{z} to \co{x},
but in the latter case there is.\footnote{
	Thanks to Peter Sewell for pointing out this possibility.}
(According to the current version of the standard, conflicting side effects
in unsequenced expressions constitute undefined behavior.
However there are proposals to make them defined in both
C++ and C,\footnote{
	\url{https://www.open-std.org/jtc1/sc22/wg21/docs/papers/2016/p0145r3.pdf}
	\url{https://www.open-std.org/jtc1/sc22/wg14/www/docs/n3203.htm}}
which would render this theoretical possibility quite real.)

The abstract executions we use will be fully specified.
This means that all the missing information must be supplied:
the implementation-defined characteristics, the selections for the
nondeterministic pathways, and most notably, for each load, the store
from which it reads and the value of the load.
We ignore issues of signal handlers and I/O;
in any case our litmus-test programs don't use them.
The totality of this information---along with the program's source
code, of course---determines within each thread a unique, linearly
ordered series of steps to be carried out by the abstract machine.
(However, there is no ordering relation between steps carried out
in different threads.  Even if a load in one thread reads from a
store in another thread, we don't require the store to come before
the load.)

With the compiler-based implementations we are considering,
the choices for the nondeterministic pathways are ``frozen'' into the
machine-code executable file and thus are completely determined
at runtime.
A consequence of this is that if two abstract executions of the same
thread under the same implementation agree on the values obtained by
the load operations during their first $N$ steps then they will agree
in every respect during those steps, although they may diverge later.

\subsection{Hardware Executions}
\label{sec:Hardware Executions}

When a given computer executes the machine code in a file, the outcome
has historically been much more well defined than the executions of the
C++ abstract machine.
The hardware's behavior is typically specified with great precision by
the designer or manufacturer, and there are formal, executable memory models
describing exactly what patterns of loads and stores can occur.
Thus, leaving aside questions of asynchronous interrupts and system
calls, the behavior of a CPU executing a particular thread within a
program is entirely determined by the values obtained by the various
load-from-memory instructions.

For this reason, the hardware executions we use will comprise (along
with machine code being run) the computer architecture and for each
load instruction, the store instruction from which it reads and the
value obtained.
At this level, the fact that the original program was in C++ is
irrelevant; the same concepts apply to the execution of a program in
any compiled language.

A computer may execute the instructions in a thread out of order.
The architecture specifies the extent to which this may happen, and it
also specifies circumstances under which some pairs of instructions
must be executed in order.
Nevertheless, we will consider an execution to be determined by the
values obtained by its loads.
As with abstract executions, if two hardware executions of the same
thread on the same type of computer agree on the values obtained by
the load instructions during their first $N$ steps then they will
agree in every respect during those steps, although they may diverge
later.

\subsection{Relation Between Abstract and Hardware Executions}
\label{sec:Relation Between Abstract and Hardware Executions}

The C++ standard requires that for any valid implementation, when a
program runs its observable behavior must be the same as that of some
abstract execution of the source code given the same input (in the
absence of any abstract executions containing undefined behavior).\footnote{
	This requirement is the standard's ``as-if'' rule.}
This means:\begin{itemize}
\item	The program's output must be the same as that of the abstract
	execution.
\item	Volatile accesses ``are evaluated strictly according to the
	rules of the abstract machine'' [4.1.2p6.1 (\co{intro.abstract})].
\item	(There is a condition on the interleaving of input and output,
	which does not matter for our purposes.)
\end{itemize}
We will say that the abstract execution is \emph{realized by} the
hardware execution.

Under any particular implementation,
a single program can have many different abstract executions,
varying in their decisions about which store each load reads from
and thus the value obtained.
It's worth noting, however, that not all the possible abstract executions
of a program need be realizable by the machine-code executable file
produced by that implementation.
In fact, we will see that \emph{none} of the possible OOTA executions
allowed by the strict C++ abstract machine will ever be realized
by the executables produced by many compilers,

Exactly what the standard's restriction on volatile accesses means
isn't entirely clear.
The handling of volatiles, as understood by compiler developers, has
been described as more folklore or a gentlemen's agreement than
anything else.
We will take it to imply that a compiler may not
invent, omit, or merge volatile accesses, in a sense to be discussed below.
